---
title: "Hindcast summary - Survival and Maturation"
author: "WGNAS 2021 - Life Cycle Model Output"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  word_document:
    reference_docx: template.docx
---

```{r global_options,include=FALSE}
knitr::opts_chunk$set(echo=FALSE, warning=FALSE, message=FALSE,fig.width=5,fig.height=3,bg="transparent", dpi = 400)
```

```{r}
# Manual table of contents, sorry if you want to change things
list_content = c("Post-smolt survival",
                 "------Complex Scale",
                 "------North America",
                 "------Northern Europe",
                 "------Southern Europe",
                 "------Correlation Matrix",
                 "------Within and between complex correlations",
                 "Probability of maturing as 1SW",
                 "------Complex Scale",
                 "------North America",
                 "------Northern Europe",
                 "------Southern Europe",
                 "------Correlation Matrix",
                 "------Within and between complex correlations")
list_page = c(2,2,2,3,3,4,5,6,6,6,7,7,8,9)
content_df <- data.frame("Content" = list_content,
                         "Page" = list_page)
nlines <- nrow(content_df)
flextable::flextable(content_df, cwidth = c(4,2), cheight = 0.2) %>%
  flextable::theme_zebra() %>%
  # flextable::fontsize(., size = 11) %>% 
  flextable::padding(., padding.top = 0, padding.bottom = 0)

```

\newpage

# Post-smolt Survival

## Complex scale

```{r,results = 'asis'}
g <- readRDS("../../temp_docx/hindcast_survival_summary_logit.rds")
grid.arrange(g)
```

## Post-smolt survival - North America

```{r,results = 'asis'}
g <- readRDS("../../temp_docx/hindcast_survival_su_North America.rds")
grid.arrange(g)
```

## Post-smolt survival - Northern Europe

```{r,results = 'asis'}
g <- readRDS("../../temp_docx/hindcast_survival_su_Northern Europe.rds")
grid.arrange(g)
```

## Post-smolt survival - Southern Europe

```{r,results = 'asis'}
g <- readRDS("../../temp_docx/hindcast_survival_su_Southern Europe.rds")
grid.arrange(g)
```

## Post-smolt survival - Correlation Matrix (logit scale)

```{r,results = 'asis',fig.width = 6, fig.height = 6}
g <- readRDS("../../temp_docx/hindcast_survival_matrix.rds")
grid.arrange(g)
```

## Post-smolt survival - between and within complex correlation (logit scale)

```{r,results = 'asis'}
g <- readRDS("../../temp_docx/hindcast_survival_boxplot.rds")
grid.arrange(g)
```

\newpage

# Prob. of maturing as 1SW

## Complex scale

```{r,results = 'asis'}
g <- readRDS("../../temp_docx/hindcast_maturation_summary_logit.rds")
grid.arrange(g)
```

## Prob. of maturing as 1SW - North America

```{r,results = 'asis'}
g <- readRDS("../../temp_docx/hindcast_maturation_su_North America.rds")
grid.arrange(g)
```

## Prob. of maturing as 1SW - Northern Europe

```{r,results = 'asis'}
g <- readRDS("../../temp_docx/hindcast_maturation_su_Northern Europe.rds")
grid.arrange(g)
```

## Prob. of maturing as 1SW - Southern Europe

```{r,results = 'asis'}
g <- readRDS("../../temp_docx/hindcast_maturation_su_Southern Europe.rds")
grid.arrange(g)
```

## Prob. of maturing as 1SW - Correlation Matrix (logit scale)

```{r,results = 'asis',fig.width = 6, fig.height = 6}
g <- readRDS("../../temp_docx/hindcast_maturation_matrix.rds")
grid.arrange(g)
```

## Prob. of maturing as 1SW - between and within complex correlation (logit scale)

```{r,results = 'asis'}
g <- readRDS("../../temp_docx/hindcast_maturation_boxplot.rds")
grid.arrange(g)
```
